Estimating the dynamics of Ghanaian mangrove cover using phenological metrics derived from synoptic Landsat observations

### 2022-04-20 manuscript-Ghana/manuscript-Ghana.tex
Big updates

---
title: Estimating the dynamics of Ghanaian mangrove cover using phenological metrics derived from synoptic Landsat observations
author:
  - name: Issoufou Liman Harou
    email: issoufoul@gmail.com
    affiliation: ICRAFGOV,KUENV
    correspondingauthor: true
  - name: Julliet Inyele 
    email: inyelejuli@gmail.com
    affiliation: KUGEO
  - name: Peter Minang  
    email: A.Minang@cgiar.org
    affiliation: ICRAFGOV
  - name: Lalisa Duguma
    email: l.duguma@cgiar.org
    affiliation: ICRAFGOV
address:
  - code: ICRAFGOV
    address: "Governance and Landscapes, World Agroforestry (ICRAF), UN Avenue, Gigiri, Nairobi, Kenya, P.O. Box: 30677-00100"
  - code: KUENV
    address: "Department of Environmental Sciences, Kenyatta University, Kenya Drive, Nairobi, Kenya, P.O. Box: 43844-00100"
  - code: KUGEO
    address: "Department of Geography, Kenyatta University, Kenya Drive, Nairobi, Kenya, P.O. Box: 43844-00100"
abstract: |
  Remote sensing analysis and narratives from Ghana (West Africa) reported evidence of mangrove degradation across the southern regions of the country. Despite the importance of mangrove forests, the coverage of earth observation satellites and the recent advances in computing power for processing spatial data, spatially-explicit estimates of mangrove vegetation cover are lacking. In these regions, optical sensors onboard Landsat satellites provide a great deal of historical remote sensing data for studying land surface dynamics.  While these sensors provide reasonable spatial, temporal and spectral resolutions for studying the dynamics of mangrove vegetation cover, a number of remote sensing artifacts make it difficult to derive a complete land cover map.  In particular, the persistence of cloud cover limits the number of usable images. To understand the dynamics of mangrove forests, we identified 17 land cover classes, harmonized the Landsat scenes across sensors, used time series analysis to improve cloud detection and recover the relevant signal encoding the temporal signature of land cover. Based on the seasonal behavior of the land cover classes, the persistence of cloud cover limiting the number of usable images and the temporal resolution of Landsat satellite, we mosaiced the images within 30 days of their acquisition dates. Using the fitted image time series, we computed unsupervised classifications which we used along with ground control polygons to generate training and validation data for land cover classifications and change detection. The analysis of the root mean square error indicates that the accuracy of the fitted image time series increases with band wavelength and land cover seasonality. The accuracy assessment indicates an overall accuracy between 0.94 and 0.95 and a Kappa between 0.93 and 0.95 for the different land cover classifications. Mangrove forests have declined in some areas, whereas they have increased in others. Mangrove area, which has decreased between 2000 and 2010, has increased between 2010 and 2020. The gains experienced between 2010 and 2020, however, did not compensate for the lost the land cover experienced between 2000 and 2010. We identified four top most important land cover involved in mangroves conversion including forests with a high rate, croplands and natural vegetation mosaics and tree plantations with moderate rate and wetlands with a low rate. Better mangrove restoration and conservation are required to fully reverse the trend of mangrove decline.
keywords: 
  - Cloud Mask
  - Google Earth Engine
  - Landsat Data
  - Land Cover Classification
  - Mangrove
journal: "Remote Sensing of Environment"
date: "`r Sys.Date()`"
# classoption: preprint, 3p, authoryear
classoption: 12pt,oneside,preprint,3p,authoryear,times
bibliography: bibliography.bib
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/remote-sensing-of-environment
csl: remote-sensing-of-environment.csl
output: 
  rticles::elsevier_article:
    latex_engine: xelatex
    keep_tex: true
    citation_package: natbib
link-citations: yes
linkcolor: blue
header-includes:
  - \input{includes.tex}
note: Replication files are available on the author's Github account (http://github.com/Issoufou-Liman).
---

```{r,include=FALSE}
options(tinytex.engine='xelatex')
pdf.options(useDingbats = TRUE)
```

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo=TRUE,out.extra='',fig.align='center',fig.pos='!htbp',cache=TRUE,echo=FALSE,warning=FALSE,message=FALSE)
```

```{r Setting-the-Scene, echo=FALSE, cache=FALSE, include=FALSE}
set.seed(123)

if (!require("remotes")) {
  install.packages("remotes")
}
if (!require("rgee")) {
  remotes::install_github("r-spatial/rgee")
}
if(!require("tinytex")) {
  install.packages("tinytex")
  tinytex::install_tinytex()
}
## loading required packages ####
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  googledrive,
  googleCloudStorageR,
  rgee,
  tidyverse,
  grid,
  gridExtra,
  tiff,
  rgdal,
  raster,
  sf,
  ggspatial,
  kableExtra
)
if (!require("gdalUtils")) install.packages("gdalUtils")

ksource <- function(x, ...) {
  source(purl(x, output = tempfile()), ...)
}

## Should local files with the same name be overwritten?
overwrite_file = FALSE

ee_Initialize(user = 'limanissoufou', drive = TRUE, gcs = FALSE)
# ee_Initialize(email = params$email, drive = params$drive, gcs = params$gcs)

asset_home = ee_get_assethome()

source('source_files/ee/01_functions.R')

if(!drive_folder_exists("Mangrove_Ghana_manuscript")){
  drive_mkdir("Mangrove_Ghana_manuscript")
}
ee_manage_create(paste0(asset_home, '/Mangrove_Ghana_manuscript'), asset_type = "Folder", quiet = FALSE)

asset_path <- paste0(asset_home, '/Mangrove_Ghana_manuscript')

dir.create(file.path('source_files'))
dir.create(file.path('source_files/ee'))
dir.create(file.path('source_files/r'))
dir.create(file.path('ee_output'))
dir.create(file.path('ee_output/Ghana_waters_polygons'))
dir.create(file.path('data_files'))
dir.create(file.path('figures'))
dir.create(file.path('figures/tifs'))
dir.create(file.path('figures/pngs'))

source('source_files/ee/02_commonSpecs.R')
map(list('.shp', '.shx', 'prj', 'dbf', 'cpg'), function(i){
  ee_to_drive_to_local (
    ee_object = ee$FeatureCollection(list(ee$Feature(roiGhana))),
    drive_description = '~/Mangrove_Ghana_manuscript/roiGhana', 
    drive_folder = 'Mangrove_Ghana_manuscript', 
    region = roiGhana, 
    scale = AgScale,
    maxPixels = 1e13, 
    local_folder  = 'ee_output/roiGhana',
    overwrite = overwrite_file,
    ee_type = 'table',
    file_extension = i
  )
})
```

```{r ee-compute-02-referenceData,include=FALSE,echo=FALSE,cache=FALSE}
assetId <- paste0(asset_path, "/Ghana_field_data")
if(!asset_exists("Ghana_field_data", asset_path)){
  source("source_files/ee/03_referenceData.R")

  ## Unsupervised class identification and matching

  # The following images are made available only. This is for 2 raisins:
  # To ensure reproducibility
  # They are time consuming in computation
  unsupervised_2000 = ee$Image('users/Liman/210311_unsupervised2000_Ghana');
  unsupervised_2010 = ee$Image('users/Liman/210311_unsupervised2010_Ghana');
  unsupervised_2018 = ee$Image('users/Liman/210311_unsupervised2018_Ghana');
  numPoints = 700
  # Class identification and matching
  past = 11; p10 = 11; przt = 11;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

  waterBodies = matches$
    stratifiedSample(
      numPoints = 250, 
      region = waterZone, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),
      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Fresh water')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 0)
      );
    }));

  past = 11; p10 = 11; przt = 11;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

  waterBodies_2 = matches$
    stratifiedSample(
      numPoints = 100, 
      region = waterZone_2, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),
      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Fresh water')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 0)
      );
    }));

  past = 11; p10 = 11; przt = 11;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

  seaWater_1 = matches$
    stratifiedSample(
      numPoints = numPoints*4/8, 
      region = seaWaterZone_1$merge(seaWaterZone_2), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Non-fresh water')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 0)
      );
    }));

  past = 9; p10 = 10; przt = 9;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')
  closeForests_1 = matches$
    stratifiedSample(
      numPoints = numPoints, 
      region = closeForestZone_1$merge(closeForestZone_2), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Close forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 1)
      );
    }));

  past = 0; p10 = 0; przt = 0;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

  openForests_1 = matches$
    stratifiedSample(
      numPoints = numPoints*2/5, 
      region = openForestZone_1, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 2)
      );
    }));

  openForests_2 = matches$
    stratifiedSample(
      numPoints = numPoints*2/5, 
      region = openForestZone_2, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 2)
      );
    }));

  openForests_3 = matches$
    stratifiedSample(
      numPoints = numPoints*1/5, 
      region = openForestZone_3, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 2)
      );
    }));

  past = 6; p10 = 6; przt = 15;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

  woodySavannas_1 = matches$
    stratifiedSample(
      numPoints = numPoints, 
      region = woodySavannaZone_1$merge(woodySavannaZone_2), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Woody savannas')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 3)
      );
    }));

  past = 3; p10 = 2; przt = 16;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');

  savannas_1 = matches$
    stratifiedSample(
      numPoints = numPoints, 
      region = savannaZone_1$merge(savannaZone_2), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Savannas')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 4)
      );
    }));

  past = 2; p10 = 2; przt = 2;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');
  closedShrubs_1 = matches$
    stratifiedSample(
      numPoints = numPoints, 
      region = closedShrubsZone_1$merge(closedShrubsZone_2)$merge(closedShrubsZone_3), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Close shrublands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 5)
      );
    }));

  past = 5; p10 = 5; przt = 1;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');
  openShrubs_1 = matches$
    stratifiedSample(
      numPoints = numPoints, 
      region = openShrubsZone_1$merge(openShrubsZone_2)$merge(openShrubsZone_3), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open Shrublands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 6)
      );
    }));

  past = 16; p10 = 16; przt = 2;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');
  grasslands_1 = matches$
    stratifiedSample(
      numPoints = numPoints, 
      region = grasslandZone_1$merge(grasslandZone_2), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Grasslands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 7)
      );
    }));

    past = 2; p10 = 6; przt = 15;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');
  wetlands_1 = matches$
    stratifiedSample(
      numPoints = numPoints*1/5, 
      region = wetlandsZone_1,
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Wetlands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 8)
      );
    }));

  wetlands_2 = matches$
    stratifiedSample(
      numPoints = numPoints*2/5, 
      region = wetlandsZone_2$merge(wetlandsZone_3), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Wetlands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 8)
      );
    }));

  wetlands_3 = matches$
    stratifiedSample(
      numPoints = numPoints*2/5, 
      region = wetlandsZone_3, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Wetlands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 8)
      );
    }));

  past = 8; p10 = 8; przt = 8;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');
  mosaic_1 = matches$
    stratifiedSample(
      numPoints = numPoints, 
      region = mosaicZone_1$merge(mosaicZone_2)$merge(mosaicZone_3), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Croplands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 9)
      );
    }));

  past = 8; p10 = 8; przt = 8;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');
  builtUp_1 = matches$
    stratifiedSample(
      numPoints = numPoints*3/5, 
      region = builtUpZone_1, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Built-up')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 10)
      );
    }));

  builtUp_2 = matches$
    stratifiedSample(
      numPoints = numPoints*2/5, 
      region = builtUpZone_2, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Built-up')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 10)
      );
    }));

  past = 5; p10 = 1; przt = 5;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');
  croplandBimodal_1 = matches$
    stratifiedSample(
      numPoints = numPoints*2.5/5, 
      region = croplandBimodalZone_1$merge(croplandBimodalZone_2)$merge(croplandBimodalZone_3)$merge(croplandBimodalZone_4)$merge(croplandBimodalZone_5), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),
      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Mosaic')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 11)
      );
    }));

  past = 16; p10 = 16; przt = 5;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');
  croplandUnimodal_1 = matches$
    stratifiedSample(
      numPoints = numPoints*2.5/5, 
      region = croplandUnimodalZone_1$merge(croplandUnimodalZone_2), 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Mosaic')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 11)
      );
    }));

>>>>

\documentclass[12pt,oneside,preprint,3p,authoryear,times]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

  \journal{Remote Sensing of Environment} % Sets Journal name

\usepackage{lineno} % add
  \linenumbers % turns line numbering on

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Estimating the dynamics of Ghanaian mangrove cover using phenological metrics derived from synoptic Landsat observations},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\input{includes.tex}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\begin{frontmatter}

  \title{Estimating the dynamics of Ghanaian mangrove cover using
phenological metrics derived from synoptic Landsat observations}
    \author[ICRAFGOV,KUENV]{Issoufou Liman Harou%
  \corref{cor1}%
  }
   \ead{issoufoul@gmail.com} 
    \author[KUGEO]{Julliet Inyele%
  %
  }
   \ead{inyelejuli@gmail.com} 
    \author[ICRAFGOV]{Peter Minang%
  %
  }
   \ead{A.Minang@cgiar.org} 
    \author[ICRAFGOV]{Lalisa Duguma%
  %
  }
   \ead{l.duguma@cgiar.org} 
      \affiliation[ICRAFGOV]{Governance and Landscapes, World
Agroforestry (ICRAF), UN Avenue, Gigiri, Nairobi, Kenya, P.O. Box:
30677-00100}
    \affiliation[KUENV]{Department of Environmental Sciences, Kenyatta
University, Kenya Drive, Nairobi, Kenya, P.O. Box: 43844-00100}
    \affiliation[KUGEO]{Department of Geography, Kenyatta University,
Kenya Drive, Nairobi, Kenya, P.O. Box: 43844-00100}
    \cortext[cor1]{Corresponding author}

  \begin{abstract}
  Remote sensing analysis and narratives from Ghana (West Africa)
  reported evidence of mangrove degradation across the southern regions
  of the country. Despite the importance of mangrove forests, the
  coverage of earth observation satellites and the recent advances in
  computing power for processing spatial data, spatially-explicit
  estimates of mangrove vegetation cover are lacking. In these regions,
  optical sensors onboard Landsat satellites provide a great deal of
  historical remote sensing data for studying land surface dynamics.
  While these sensors provide reasonable spatial, temporal and spectral
  resolutions for studying the dynamics of mangrove vegetation cover, a
  number of remote sensing artifacts make it difficult to derive a
  complete land cover map. In particular, the persistence of cloud cover
  limits the number of usable images. To understand the dynamics of
  mangrove forests, we identified 17 land cover classes, harmonized the
  Landsat scenes across sensors, used time series analysis to improve
  cloud detection and recover the relevant signal encoding the temporal
  signature of land cover. Based on the seasonal behavior of the land
  cover classes, the persistence of cloud cover limiting the number of
  usable images and the temporal resolution of Landsat satellite, we
  mosaiced the images within 30 days of their acquisition dates. Using
  the fitted image time series, we computed unsupervised classifications
  which we used along with ground control polygons to generate training
  and validation data for land cover classifications and change
  detection. The analysis of the root mean square error indicates that
  the accuracy of the fitted image time series increases with band
  wavelength and land cover seasonality. The accuracy assessment
  indicates an overall accuracy between 0.94 and 0.95 and a Kappa
  between 0.93 and 0.95 for the different land cover classifications.
  Mangrove forests have declined in some areas, whereas they have
  increased in others. Mangrove area, which has decreased between 2000
  and 2010, has increased between 2010 and 2020. The gains experienced
  between 2010 and 2020, however, did not compensate for the lost the
  land cover experienced between 2000 and 2010. We identified four top
  most important land cover involved in mangroves conversion including
  forests with a high rate, croplands and natural vegetation mosaics and
  tree plantations with moderate rate and wetlands with a low rate.
  Better mangrove restoration and conservation are required to fully
  reverse the trend of mangrove decline.
  \end{abstract}
    \begin{keyword}
    Cloud Mask \sep Google Earth Engine \sep Landsat Data \sep Land
Cover Classification \sep 
    Mangrove
  \end{keyword}

 \end{frontmatter}

\hypertarget{software-and-data-availability}{%
\section*{Software and data
availability}\label{software-and-data-availability}}
\addcontentsline{toc}{section}{Software and data availability}

We conducted the data analysis in Google Earth Engine cloud computing
platform \citep{Gorelick-et-al-2017} and R programming language
\citep{R-Core-Team-2021}. These were interfaced using rgee R package
\citep{Aybar-et-al-2020} which we used as bridge for throughput between
Google Earth engine and R. Both Google Earth engine and R are freely
accessible (see \url{https://cran.r-project.org/index.html} for R and
\url{https://code.earthengine.google.com/} for Google Earth Engine). We
automated the entire process, including the installation of the required
R packages; data processing; data streaming between Google Earth Engine
and R; and results visualization, to provide a complete reproducible
workflow. The replication files along with further information for
reproducing the work are available from
\url{https://github.com/Issoufou-Liman/mangrove-RSE}.

\hypertarget{definition-of-key-terms}{%
\section*{Definition of key terms}\label{definition-of-key-terms}}
\addcontentsline{toc}{section}{Definition of key terms}

In this section, we define several terms which may lead to confusion.
Remote sensing artifacts regroup all atmospheric (e.g., such as cloud,
cloud shadow, haze, aerosol scattering) and technical factors (e.g.,
scan line corrector failure in Landsat 7 Enhanced Thematic Mapper Plus)
that would potentially lead to unrealistic pixel value estimates. The
term ``clear'' (as in clear pixels or clear observation) refers to
pixels exempt of remote sensing artifacts as estimated by Fmask
algorithm. We used the term ``cloudy'' (as in cloudy observations or
cloudy pixels) to denote pixel value identified by Fmask algorithm as
affected by remote sensing artifacts. These are different from ``noisy''
(as in noisy observations). Noisy pixels are clear pixels whose values
are not reasonably within the range of valid values. These are mostly
cloudy pixels not identified by Fmask algorithm. We used the term
``signal'' to denote the smooth temporal distribution of pixel values
that describes the average phenological profile at the pixel. This is
smooth data describing the phenology of LULC and the information of
interest as opposed to ``noise'' which denotes the random fluctuations
that obscures the signal. We used the term ``overall study period'' to
denote the period 2000 to 2020, the term ``previous decade'' to denote
the period 2000 to 2010, and the term ``last decade'' to denote the
period 2010 to 2020.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Mangroves are coastal forests that grow where ocean water, freshwater,
and land meet \citep{CILSS-2016}. They are highly productive forest
ecosystems found in the intertidal zone of tropical and subtropical
coasts
\citep{Spalding-et-al-1997, Spalding-et-al-2010, Wang-et-al-2019}.
Mangrove trees develop adaptation mechanisms that enable them to survive
in brackish water conditions
\citep{Spalding-et-al-1997, Spalding-et-al-2010}. They play important
roles in regulating coastal processes, in providing various ecosystem
services to the socio-ecological systems in the coastal regions, or in
underpinning traditional customs for many communities
\citep{Ajonina-et-al-2013, Levy-et-al-2015, UNEP-2007, Wilkie-and-Fortuna-2003}.

Recent studies \citep{Rubin-et-al-1999, Wilkie-and-Fortuna-2003}
estimated mangroves to cover between 14 and 17 million hectares of
tropical coasts. Around 3.2 million ha of these (19\% of global
coverage) are found in Africa \citep{Ajonina-et-al-2013}, with West
Africa hosting 70\% of the total mangrove areas of the continent
\citep{UNEP-2007}. In Ghana, the area covered by mangroves is estimated
to range between 137 to 140 km2 \citep{Armah-et-al-2009, UNEP-2007}.
Mangroves provide a substantial contribution to coastal fisheries, which
contribute to around \$400 million per year to the regional economy of
West Africa \citep{USGS-EROS-2016}. In Ghana, this value is estimated at
\$6 million per year \citep{Armah-et-al-2009}.

Despite the importance of mangroves, some studies reported that mangrove
vegetation cover has been alarmingly declining in many areas of Ghana
for the last few decades \citep{Asante-et-al-2017}.
\citet{Armah-et-al-2009} estimated this decline to range between 20 and
30\% across the coast of Atlantic Ocean over the past 3 decades.
\citet{FAO-2007} reported a mangrove vegetation loss of 0.5 million ha
(13.8 \% of the total mangrove areas) in 25 years. According to
\citet{Rubin-et-al-1999}, almost 2/3 of the Volta estuary mangroves have
been lost since 1973, leaving only a few patches within a buffer of 15
km. \citet{Darkwa-and-Smardon-2010} noticed an increased cutting of
mangroves vegetation in Fosu Lagoon. \citet{FAO-2007} estimated the
decline of mangrove areas in Ghana at 24\% between 1980 to 2005. In
other areas, however, other studies reported an appreciable increase in
mangrove vegetation cover following restoration efforts by the locals.
\citet{Awo-et-al-2014} reported a noticeable increase in mangrove cover
between 1986 and 2002 in Anyanui (Keta Lagoon). \citet{Feka-2015} noted
that despite the reduction in mangrove area in Ghana, an estimated area
of 68 ha of mangroves have been restored from 1980 to 2006.

It becomes clear from the available information that a country-wide
assessment of mangrove vegetation in Ghana is needed. A holistic
assessment of all major Land Use Land Cover (LULC) would help to
understand the dynamics of mangrove vegetation and guide mangrove
conservation policy in the study area. This would help identify areas of
increase and decrease in mangrove vegetation, understand the trend in
LULC change, identify areas that require immediate action and those
areas where policy interventions can achieve the greatest impact. To our
knowledge, country-wide assessments of mangrove vegetation that deliver
this crucial information for mangrove conservation are lacking. The most
recent and comprehensive ones (e.g., \citet{CILSS-2016};
\citet{Spalding-et-al-1997}; \citet{Spalding-et-al-2010}) have either
coarse spatial resolution or have important information gap to be
suitable for this purpose. This information gap is mainly due to the
prevalence of cloud cover \citep{Ashiagbor-et-al-2021} that makes it
virtually impossible to derive a complete land cover map from a single
acquisition \citep{Rubin-et-al-1999}. Clouds deserved a standalone class
in \citet{CILSS-2016} because none of the best available image was
cloud-free enough to allow a comprehensive post-classification
interpolation. Based on our analysis of clouds identified by Fmask
algorithm, at least half of the available Landsat images acquired
between 2000 and 2020 presents missing observations over the coastal
regions of Ghana. There is a need for approaches that could
comprehensively account for these gaps when working with optical remote
sensing data which are currently the best available source of historical
images in the coastal region of Ghana.

The use of continuous time series of satellite images, which provide
consistently repeated observations of the earth's surface, can help
improve the accuracy of mangroves estimates in Ghana
\citep{Lambin-and-Strahlers-1994, Verbesselt-et-al-2010, Zhu-and-Woodcock-2014}.
However, differences due to remote sensing artifacts in these
multitemporal images may cause error in comparative image analysis.
Change detection analysis, therefore, should involve comparable data
(i.e., similar sensors, similar radiometric and spatial resolutions,
similar viewing geometries, similar image anniversaries). The
reliability of comparative image analysis for change detection such as
post-classification, temporal image differencing, temporal image
ratioing, depends on various environmental and atmospheric factors and
the accuracy of their resulting change detection analysis depends on the
accuracy of the input data. Error in the input images is likely to have
compounding effects in the change detection results. These approach,
however, are intuitive, simple to implement, and provide quantitative
change estimates across a variety of LULC. Yet, for these conventional
methods to effectively support any operational application, data input
should be comparable across the spatial and temporal domains of the
analysis \citep{Lillesand-et-al-2015}. Multitemporal image analysis
should account for potential sources of uncertainty such as imperfection
of cloud masking algorithms, sensor differences, and other ephemeral
changes induced by erratic events that could introduce error in the
classifications \citep{Zhu-and-Woodcock-2014}. Multi-date images should
be of the same period to account for differences in sun angle and
temporal variability in LULC. Post-classifications for change detection
should account for all major LULC while considering reference data
(i.e., training and validation data) over areas of persistent LULC.
Besides the requirements for training and validation for LULC
classifications, it is important to ensure the comparability of the LULC
classifications \citep{Lillesand-et-al-2015, Zhu-and-Woodcock-2014}.
This can be achieved by either using a single classifier that is
consistent with all classifications considered or ensuring that
throughputs and accuracies are comparable when using multiple
classifiers. These basics requirements have strong implications for the
accuracy of change detection analysis that rely on multitemporal image
classification. These requirements are often difficult to satisfy due to
the lack of ideal satellite images. While Landsat collections are
currently the best publicly available source of data that provide a
reasonable and consistent synoptic estimate of land surface state within
acceptable spatial resolution for studying changes in mangrove
vegetation, it is virtually impossible to obtain the ideal pair of
Landsat images in the coastal regions of Ghana. In these areas, the
persistence of thick cloud layers reduces the number of usable images
\citep{Rubin-et-al-1999}. Consequently, these images are not always
available at the nominal revisit period of 16 days for every pixel.
